---
title: "R Learning Journey"
author: "Timothy Borghouts"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

# Table of Contents

1.  [Introduction](#introduction)
2.  [Variables & Data Types](#variables-data-types)
3.  [Data Structures](#data_structures)
4.  [Basic Operations](#basic_operations)

---

# 1. Introduction {#introduction}

Hello! I'm Timothy Borghouts, a 20-year-old student living in Breda. I'm currently studying at Avans University of Applied Sciences.

I’ve been learning Python as part of my study, along with many other programming languages like java, C# javascript and others. Despite that, I’ve always been curious about R and have decided to start learning it on my own. This document will track my journey as I explore R, from basic variables to more advanced data visualization.

This is my R Learning Notebook, where I document everything I’ve learned throughout my journey with R.

---

# 2. Variables & Data Types {#variables-data-types}

Data types in R are the basic building blocks for all variables and data structures. They include numeric, character, logical, and complex types.

## 2.1. Numeric Variables
Numeric variables store numbers, including both whole numbers (integers) and decimal values (floating-point numbers).

```{r}         
numeric <- 10 # Numeric
integer <- 20L # Integer 
float <- 12.34 # Float

# Printing numeric variables
cat("Numeric:", numeric, "\nInteger:", integer, "\nFloat:", float)
```

## 2.2. Character Variables
Character variables represent text or string data, enclosed in quotes.

```{r}         
character <- "Timothy"  # Character
sentence <- "Timothy <3 Malou" # longer text
multiline <- "
Malou is 
as slow as 
a snail" # multiline

# Printing character variables
cat("Character: ", character, "\nSentence:", sentence, "\nMultiline:", multiline)
```

## 2.3. Logical Variables
Logical variables represent boolean values, typically TRUE or FALSE, used in decision-making and comparisons.

```{r}         
is_true <- TRUE  # Logical
is_false <- FALSE 
true <- T # True
false <- F # False

# Printing logical variables
cat(is_true, is_false, true, false)
```

## 2.4 Complex Variables
Complex variables represents complex numbers with both real and imaginary components (e.g., 2 + 3i).

```{r}        
complex <- 17i # Complex
more_complex <- 2 + 3i # Complex
# most_complex <- "Malou"

# Printing complex variables
cat(complex, "\n", more_complex)
```

---

# 3. Data structures {#data_structures}

Data structures in R are used to organize and store data efficiently. These include vectors, matrices, arrays, lists, and data frames, which allow for more complex data manipulation and analysis.

## 3.1. Vectors
A sequence of elements of the same type, such as numbers, characters, or logical values.

```{r}         
numbers <- c(1, 3, 13, 4, 9) # Vector with numerical values
numbers_in_sequence <- 1:10 # Vector with numerical sequence 1 to 10
characters <- c("Ikea", "Bed", "Sleep") # Vector of characters
logicals <- c(TRUE, FALSE, TRUE, FALSE) # Vector of logical values

# Printing vectors
cat(numbers, "\n", numbers_in_sequence, "\n", characters, "\n", logicals)
```

## 3.2. Lists
A flexible structure that can contain multiple types of elements, including numbers, text, and even other lists.

```{r}
character_list <- list("R", "C#", "Java", "Python") #List with character values
numerical_list <- list(1, 24, 9, 183, 2) #List with numerical values
named_list <- list(name = "Timothy", age = 20, length = 1.83) #List with named elements
list_of_lists <- list(character_list, numerical_list, named_list) # List containing other lists

# Printing lists
cat("Character List: ", toString(character_list), "\nNumerical List: ", toString(numerical_list), "\nNamed List:     ", toString(named_list))
```

## 3.3. Matrices
A two-dimensional arrangement of elements of the same type, organized into rows and columns.

```{r}   
# Numeric matrix (3x3)
numeric_matrix <- matrix(1:9, nrow = 3, ncol = 3) 
# Character (string) matrix (2x3)
string_matrix <- matrix(c("A", "B", "C", "D", "E", "F"), nrow = 2, ncol = 3) 
# Matrix with row/column names
complex_matrix <- matrix(c(10, 20, 30, 40, 50, 60, 70, 80, 90), nrow = 3, ncol = 3) 
rownames(complex_matrix) <- c("Row1", "Row2", "Row3")
colnames(complex_matrix) <- c("Col1", "Col2", "Col3")


# Printing Matrices
cat("Numeric Matrix:\n")
print(paste(numeric_matrix))
cat("String Matrix:\n")
print(string_matrix)
cat("Matrix with Row and Column Names:\n")
print(complex_matrix)
```


## 3.4. Arrays
A multi-dimensional extension of matrices, capable of storing data in multiple layers.

```{r}        
# 1D array (Just like a vector)
one_d_array <- array(1:6)  


# 2D array with 3 rows and 3 columns (just like a matrix)
two_d_array <- array(1:9, dim = c(3, 3))  


# 3D array (3 rows, 3 columns, 3 layers)
three_d_array <- array(1:27, dim = c(3, 3, 3))  


# Printing Arrays
cat("1D Array (Vector):\n")
print(one_d_array)
cat("\n2D Array (Matrix):\n")
print(two_d_array)
cat("\n3D Array:\n")
print(three_d_array)
```


## 3.5. Data Frames
A table-like structure where each column can hold different types of data, commonly used for organizing and analyzing datasets.

```{r}         
data_frame <- data.frame( # Data frame
  Name = c("Timothy", "Malou", "Kim", "Freek"), # Vector of names
  Age = c(20, 20, 20, 19), # Vector of numbers
  Height = c(6.0, 2.3, 5.7, 6.2) # Vector of floats
)

data_frame$Name  # Access the 'Name' column
data_frame[1, 2]  # Access element at first row, second column

# Display the entire data frame
data_frame
```

## 3.6. Factor
A structure used for categorical data, storing values as predefined levels for efficient grouping.

```{r}        
factor <- factor(c("Cold", "Normal", "Malou")) # Factor

# Printing factors
factor
levels(factor)  # Get the levels of the factor
```

---

# 4. Basic Operations {#basic_operations}

Basic operations in R are fundamental for mathematical calculations and data manipulation. These operations include arithmetic, logical, and comparison operations, which allow you to perform calculations and make decisions based on data values.

## 4.1 Arithmetic Operations
These operations allow you to perform basic mathematical calculations in R.

| **Operation**        | **Operator** | **Example Input**   | **Example Output** | **Explanation** |
|----------------------|--------------|---------------------|--------------------|-----------------|
| **Addition**         | `+`          | `10 + 2`            | `12`               | Adds two values. |
| **Subtraction**      | `-`          | `9 - 3`             | `6`                | Subtracts the second value from the first. |
| **Multiplication**   | `*`          | `5 * 5`             | `25`               | Multiplies two values. |
| **Division**         | `/`          | `9 / 3`             | `3`                | Divides the first value by the second. |
| **Power**            | `^` or `**`  | `5 ^ 2` or `4 ** 2` | `25` or `16`       | Raises the first number to the power of the second number. |
| **Modulus**          | `%%`         | `10 %% 3`           | `1`                | Finds the remainder when dividing the first value by the second. |
| **Integer Division** | `%/%`        | `10 %/% 3`          | `3`                | Divides the first value by the second and returns the quotient (ignores the remainder). |


```{r}
x <- 10
y <- 3

addition <- x + y       # 13
subtraction <- x - y    # 7
multiplication <- x * y # 30
division <- x / y       # 3.33
power1 <- x ^ y         # 1000
power2 <- x ** y        # 1000
modulus <- x %% y       # 1
int_div <- x %/% y      # 3

cat("addition:", addition, "\n",
    "subtraction:", subtraction, "\n",
    "multiplication:", multiplication, "\n",
    "division:", division, "\n",
    "power1:", power1, "\n",
    "power2:", power2, "\n",
    "modulus:", modulus, "\n",
    "int_div:", int_div, "\n")
       
```

## 4.2 Logical operations
Logical operations evaluate boolean expressions (TRUE or FALSE). These operations are used for logical comparisons.

| **Operation** | **Operator** | **Example Input**   | **Example Output** | **Explanation** |
|---------------|--------------|---------------------|--------------------|-----------------|
| **AND**       | `&`          | `TRUE & FALSE`      | `FALSE`            | Returns `TRUE` only if both values are `TRUE`. |
| **OR**        | `||`         | `TRUE || FALSE`     | `TRUE`             | Returns `TRUE` if at least one value is `TRUE`. |
| **NOT**       | `!`          | `!TRUE`             | `FALSE`            | Reverses the boolean value (NOT operation). |


```{r}
a <- TRUE
b <- FALSE

and_result <- a & b  # FALSE
or_result <- a || b  # TRUE
not_result <- !a     # FALSE

cat("and_result:", and_result, "\n",
    "or_result:", or_result, "\n",
    "not_result:", not_result, "\n")
```

## 4.3 Comparison Operations
These operations allow you to compare values to evaluate conditions.

| **Operation**        | **Operator** | **Example Input**   | **Example Output** | **Explanation** |
|----------------------|--------------|---------------------|--------------------|-----------------|
| **Equal to**         | `==`         | `3 == 3`            | `TRUE`             | Returns `TRUE` if both values are equal. |
| **Not equal to**     | `!=`         | `3 != 4`            | `TRUE`             | Returns `TRUE` if the values are not equal. |
| **Greater than**     | `>`          | `5 > 2`             | `TRUE`             | Returns `TRUE` if the first value is greater than the second. |
| **Less than**        | `<`          | `5 < 2`             | `FALSE`            | Returns `TRUE` if the first value is less than the second. |
| **Greater or Equal** | `>=`         | `5 >= 5`            | `TRUE`             | Returns `TRUE` if the first value is greater than or equal to the second. |
| **Less or Equal**    | `<=`         | `3 <= 2`            | `FALSE`            | Returns `TRUE` if the first value is less than or equal to the second. |


```{r}
x <- 3
y <- 4
z <- 5

equal <- x == y   # FALSE
not_equal <- x != y # TRUE
greater_than <- z > x # TRUE
less_than <- z < y    # FALSE
greater_or_equal <- z >= x  # TRUE
less_or_equal <- x <= y    # TRUE

cat("equal:", equal, "\n",
    "not_equal:", not_equal, "\n",
    "greater_than:", greater_than, "\n",
    "less_than:", less_than, "\n",
    "greater_or_equal:", greater_or_equal, "\n",
    "less_or_equal:", less_or_equal, "\n")
```

